---
title: "Fertility and Reproductive Timing among Women in Nigeria"
subtitle: "Findings from the 2018 Demographic and Health Survey"
author: "GRAPH Courses Team"
date: "Dec 24, 2022"
output: prettydoc::html_pretty
editor_options: 
  chunk_output_type: console
---

```{r echo = F}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

```{r packages-and-plots}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Packages ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

if(!require(pacman)) install.packages("pacman")
pacman::p_load(
  janitor, # data analysis utilities
  here, # force rmds to use the project folder as working directory
  haven, # for reading in stata files
  scales, # for the comma() function, among others
  ggstatsplot, # for pie charts
  patchwork, # combining plots
  tidyverse # for everything!
)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ggplot2 palette  ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# manual color scale for use later. 
# Got the scale for this website: # https://mycolor.space/?hex=%2354B1BF&sub=1
my_palette <- c("#98d8e3","#74cddb","#54B1BF","#288D9B","#006B78","#004A57")

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Import and process data  ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# IR stands for individual recode
ir_raw <- haven::read_dta(here("data/NGIR7BFL.DTA"),
                          # col_select is used to avoid importing the entire dataset, 
                          # which is very large
                          col_select = c(v012, v022, v013, v201, v213, v024, v190, 
                                         v531, v511, v212, v149, v006, v007, v009, 
                                         v010, v025))



ir <- 
  ir_raw %>% 
  
  # rename the needed variables
  rename(age = v012, 
         state = v022,
         age_group = v013, 
         num_kids = v201, 
         curr_pregnant = v213,
         region = v024, 
         wealth_index = v190, 
         age_first_sex = v531, 
         age_first_cohabitation = v511, 
         age_first_birth = v212, 
         educ = v149, 
         mth_interview = v006, 
         yr_interview = v007, 
         mth_birth = v009, 
         yr_birth = v010, 
         urban_rural = v025) %>% 
  
  # convert from labelled data to regular factors
  haven::as_factor() %>% 
  
  # simplify education categories
  mutate(
    educ = as.character(educ),
    educ = case_when(
      educ %in% c("incomplete primary", "complete primary") ~ "Primary",
      educ %in% c("incomplete secondary", "complete secondary") ~ "Secondary",
      educ == "no education" ~ "None",
      educ == "higher" ~ "Higher",
      TRUE ~ educ),
    educ = factor(educ, 
                  levels = c("None", "Primary", "Secondary", "Higher")))
```


```{r fertility}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Fertility ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

num_women_in_survey <- nrow(ir)
num_with_child <- ir %>% filter(num_kids > 0) %>% nrow()
pct_with_child <- (100 * num_with_child/num_women_in_survey) %>% round(digits = 1)
mean_sd_kids <- 
  ir %>% summarise(mean_kids = round(mean(num_kids), digits = 2), 
                   sd_kids = round(sd(num_kids), digits = 2))

gg_col_children_born <- 
  ir %>% 
  count(num_kids) %>% 
  mutate(pct = 100 * n/sum(n),
         pct = round(pct, digits = 0), 
         pct = as.character(pct),
         pct = if_else(pct <1, "<1", pct),
         pct = paste0(pct, "%")) %>% 
  ggplot(aes(x = num_kids, y = n)) + 
  geom_col(fill = "#006B78")  +
  geom_text(aes(y = n + 400, 
                label = pct),
            size = 3.3, 
            fontface = "bold")  +
  labs(x = "Children ever born", y = "Frequency count") +  
  scale_y_continuous(expand = c(0, 0), 
                     limits = c(0, 13500)) +
  scale_x_continuous(breaks = 0:17)+
  theme_classic()
```


```{r reproductive-events}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Age at first sex, marriage and birth ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# The analyses below are only carried out on individuals aged 25 up ----
# This is because most younger (e.g. those aged 16), are not married and have not yet given birth, 
# so calculating their median age at marriage, birth is not meaningful

twentyfive_up <- 
  ir %>% 
  filter(age >= 25) %>%
  # convert factors to numbers
  # missing values, such as "don't know" become `NA`
    mutate(age_first_sex = as.numeric(as.character(age_first_sex))) %>% 
    mutate(age_first_cohabitation = as.numeric(as.character(age_first_cohabitation))) %>% 
    mutate(age_first_birth = as.numeric(as.character(age_first_birth))) 

count_women_twentyfive_up <- twentyfive_up %>% nrow() 

### First sex----
count_women_twentyfive_up_had_sex <- twentyfive_up %>% filter(!is.na(age_first_sex)) %>% nrow() 
pct_women_twentyfive_up_had_sex <- (100 * count_women_twentyfive_up_had_sex/count_women_twentyfive_up) %>% round(1)
median_first_sex <- median(twentyfive_up$age_first_sex, na.rm = T)

### First cohabitation (marriage or informal union) ----
count_women_twentyfive_up_have_cohabited <- twentyfive_up %>% filter(!is.na(age_first_cohabitation)) %>% nrow() 
pct_women_twentyfive_up_have_cohabited <- (100 * count_women_twentyfive_up_have_cohabited/count_women_twentyfive_up) %>% round(1)
median_first_cohabitation <- median(twentyfive_up$age_first_cohabitation, na.rm = T)
  
### First birth----
count_women_twentyfive_up_have_had_child <- twentyfive_up %>% filter(!is.na(age_first_birth)) %>% nrow() 
pct_women_twentyfive_up_have_had_child <- (100 * count_women_twentyfive_up_have_had_child/count_women_twentyfive_up) %>% round(1)
median_first_birth <- median(twentyfive_up$age_first_birth, na.rm = T)


### Life course events per education group----

gg_col_events_per_educ_group <- 
  twentyfive_up %>% 
  group_by(educ) %>% 
  summarise(med_age_first_sex = median(age_first_sex, na.rm = T), 
            med_age_first_cohabitation = median(age_first_cohabitation, na.rm = T), 
            med_age_first_birth = median(age_first_birth, na.rm = T)) %>% 
  pivot_longer(2:4) %>% 
  # set the levels so that they show up in a sort of chronological order on ggplot
  mutate(name = factor(name,
                       levels = c("med_age_first_sex", 
                                  "med_age_first_cohabitation", 
                                  "med_age_first_birth"))) %>% 
  
  # pass the data directly into ggplot
  ggplot() + 
  geom_col(aes(x = name, y = value, fill = educ), position = "dodge") + 
  geom_text(aes(x = name, y = value - 1.2, group = educ, label = value), 
            size = 3.8, fontface = "bold", color = "white",
            position = position_dodge(width = 0.9) # figure out dodged amount by trial and error
            ) +
  
  # use manual palette. Looks better reversed with `rev()`
  scale_fill_manual(values = rev(my_palette)) + 
  
  # Presentable x axis labels
  scale_x_discrete(labels = c("Median age \nat first sex", 
                              "Median age \nat first cohabitation", 
                              "Median age \nat first birth")) +
  
  scale_y_continuous(expand = c(0, 0), # prevent expansion of scale
                     limits = c(0, 27) # expand scale to prevent clipping of labels
                     )  +
  labs(title = "Median ages at key events by education group", 
       x = "", y = "Age (years)", fill = "Education Level")  +
  theme_classic()

### Life course events per residence (urban/rural) group----

gg_col_events_per_urban_rural_group <- 
  twentyfive_up %>% 
  group_by(urban_rural) %>% 
  summarise(med_age_first_sex = median(age_first_sex, na.rm = T), 
            med_age_first_cohabitation = median(age_first_cohabitation, na.rm = T), 
            med_age_first_birth = median(age_first_birth, na.rm = T)) %>% 
  pivot_longer(2:4) %>% 
  mutate(name = factor(name,
                       levels = c("med_age_first_sex", 
                                  "med_age_first_cohabitation", 
                                  "med_age_first_birth"))) %>% 
  mutate(urban_rural = fct_rev(urban_rural)) %>% 
  ggplot() + 
  geom_col(aes(x = name, y = value, fill = urban_rural), position = "dodge") + 
  geom_text(aes(x = name, y = value - 1.2, group = urban_rural, label = value), 
            size = 3.8, fontface = "bold", color = "white",
            position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c(my_palette[5], my_palette[3])) +
  scale_x_discrete(labels = c("Median age \nat first sex", 
                              "Median age \nat first cohabitation", 
                              "Median age \nat first birth")) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 27))  +
  labs(title = "Median ages at key events by residence type", 
       x = "", y = "Age (years)", fill = "Residence type") +
  theme_classic()

### Combine plots ----

gg_events <- patchwork::wrap_plots(gg_col_events_per_educ_group, 
                      gg_col_events_per_urban_rural_group, ncol = 1) + 
  plot_annotation(tag_levels = 'a')
```


```{r teen-fertility}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Teenage fertility ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

teen_childbearing <- 
  ir %>% 
  # Subset to teens
  filter(age_group == "15-19") %>% 
  mutate(
    # group people with children OR people who are currently pregnant into a single category
    childbearing = case_when(num_kids > 0 | curr_pregnant == "yes" ~ "Mother or pregnant",
                                  TRUE ~ "Neither mother nor pregnant"),
         # set factor level order
         childbearing = factor(childbearing, levels = c("Mother or pregnant", 
                                                        "Neither mother nor pregnant")))


num_teens <- teen_childbearing %>% nrow()
num_teens_childbearing <- teen_childbearing %>% filter(childbearing == "Mother or pregnant") %>% nrow()
pct_teens_childbearing <- (100 * num_teens_childbearing/num_teens) %>% round(digits = 1)

### Overall teen childbearing percentage pie chart ----

gg_pie_childbearing  <- 
  teen_childbearing %>%
  ggstatsplot::ggpiestats(
    x = childbearing,
    title = "Percent of teenagers (15-19) who have begun childbearing",
    results.subtitle = F # turn off statistical tests
    ) +
  theme_void() +
  labs(fill = "Childbearing status") + 
  theme(axis.line.x.bottom = element_blank(), 
        legend.position = "right") +
  scale_fill_manual(values = c("#54B1BF", "#004A57")) + 
  guides(fill = guide_legend(reverse=T)) # legend order (can be found by Googling "how do I reverse legend order in ggplot")


### Teen childbearing by region ----

teen_childbearing_by_region <- 
  teen_childbearing %>% 
  group_by(region) %>% 
  count(childbearing) %>% 
  mutate(pct = 100 * n/sum(n)) %>% 
  filter(childbearing == "Mother or pregnant")  %>% 
  ungroup() %>% 
  mutate(region = str_to_title(region)) %>% 
  mutate(region = fct_reorder(region, pct)) %>% 
  arrange(-n)

# Select regions with the highest and lowest teen childbearing rates
region_hi_teen_child_rate <- teen_childbearing_by_region %>% head(1)
region_lo_teen_child_rate <- teen_childbearing_by_region %>% tail(1)

# Plot the teen childbearing by region data
gg_col_teen_childbearing_by_region <- 
  teen_childbearing_by_region %>% 
  ggplot(aes(y = region, x = pct, fill = pct)) + 
  geom_col() + 
  scale_fill_gradient(low = "#74cddb", high = "#004A57") + 
  scale_x_continuous(expand = c(0,0)) +
  labs(title = "Teenage childbearing by region", 
       fill = "% who have \nbegun \nchildbearing", 
       x = "Percent", 
       y = "Region") +
  theme_classic() + 
  theme(axis.line.y = element_line(), 
        legend.position = "right")

# Combine plot
gg_teen_childbearing <- 
  patchwork::wrap_plots(gg_pie_childbearing, 
                        gg_col_teen_childbearing_by_region, 
                        ncol = 1) + 
  plot_annotation(tag_levels = 'a')

```

# Fertility

Across the `r comma(num_women_in_survey)` surveyed women (ages 15 to 49), `r comma(num_with_child)`, 
(`r pct_with_child`%) had at least one child. 

```{r fig.height = 4}
#| fig.cap = "Distribution of the number of children born among surveyed women."
gg_col_children_born
```

Figure 1 shows the distribution of the number of children per women. The
mean number of children ever born was `r mean_sd_kids$mean_kids`, with a standard deviation of `r mean_sd_kids$sd_kids`.

# Age at first sex, marriage and birth

Of `r comma(count_women_twentyfive_up)` women aged 25-49 in the survey,
`r comma(count_women_twentyfive_up_had_sex)`, ( `r pct_women_twentyfive_up_had_sex`% )
have had sex, `r comma(count_women_twentyfive_up_have_cohabited)`, ( `r pct_women_twentyfive_up_had_sex`% )
have entered a cohabiting relationship (marriage or informal union), and
`r comma(count_women_twentyfive_up_have_had_child)`, ( `r pct_women_twentyfive_up_have_had_child`% ) have had at least one child.

The overall median age of first sex was `r median_first_sex` years, 
while the median first cohabitation was `r median_first_cohabitation` years, 
and the median for first birth was `r median_first_birth` years.

```{r fig.height = 8}
#| fig.cap = "Median ages at first sex, cohabitation and birth by education level and residence type. 
#| Cohabitation refers to people living together, either in a formal marraige or in an informal union"
gg_events
```

Figure 2 shows median ages at key life events by education group and residence type. 
In general, women with more education had first sex, first cohabitation and first birth later than those with less education. 
Women in urban areas also experienced these events later than those living in rural areas. 

# Teenage Fertility

```{r fig.height = 6}
#| fig.cap = "Percentage of teenage women ages 15 to 19 who have begun childbearing, overall and by region. 
#| A womam who has begun childbearing has had a child or is currently pregnant."
gg_teen_childbearing
```

Figure 3 shows the percentage of teenage women who have begun childbearing, overall and by region. 
Out of `r comma(num_teens)` teenage women (15-19) surveyed, 
`r comma(num_teens_childbearing)` (`r pct_teens_childbearing` %) are mothers or pregnant with their first child. 
By region, the teenage childbearing rate ranges from `r round(region_hi_teen_child_rate$pct)`% in `r region_hi_teen_child_rate$region` 
to  `r round(region_lo_teen_child_rate$pct)`% in `r region_lo_teen_child_rate$region`.



